---
title: "Playing with mixture models"
author: "John Fee"
date: "2023-03-25"
output: bookdown::github_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
  )
```

```{r load_packages}
library(magrittr)
library(dplyr)
library(tidyr)
library(purrr)

library(ggplot2)
theme_set(cowplot::theme_cowplot())

set.seed(1)
```


# Analytic and synthetic views on mixture models

The analytical formulation for mixture models is a convex combination of densities.  That is,

\begin{align}
y \sim \sum_{g = 1}^{G}\eta_{g}f(\cdot|\theta_{g})
\end{align}

where the the $\eta_{g}$ terms sum to 1 and the parameters of the density function $f$ are allowed to vary (though this could be relaxed).  Contrast this synthetic (constructivist?) approach where for each observation $i$ in our dataset we sample the index set $g = \{1,2,\ldots,G\}$ using the weights $\eta_{g}$, and then sample from the chosen density $f_{g}$.

\begin{align}
y_{i} | z_{i} ~ f(\cdot | \theta_{z_{i}}) && \text{with } P(z_{i} = g = \eta_{g})
\end{align}


Let's use the synthetic representation to simulate what a mixture distribution could look like!

```{r normal-simulation,cache = TRUE}
#| fig.cap = "A univariate mixture of normal distributions with generating model is $f(y) \\sim .2 \\times \\mathcal{N}(1,1) + .8 \\times \\mathcal{N}(5,1)$"

sim_parameters <- data.frame(
  eta = c(.2,.8),
  mu = c(1,5),
  sigma = c(1,1)
)

sample_synthetic <- function(n,sim_parameters){
  # Get indices to use for each density function
  my_index <- sample(
    x = 1:nrow(sim_parameters),
    prob = sim_parameters$eta,
    size = n,
    replace = TRUE
    ) %>%
    data.frame(index = .) %>%
    count(index)

  # Generate # samples from each density according sampled density indices
  my_index %>%
    purrr::pmap(
      function(index,n) rnorm(n,sim_parameters$mu[index],sim_parameters$sigma[index])
    ) %>%
    purrr::flatten_dbl()
  
}

n <- 1000

data.frame(
  "synthetic" = sample_synthetic(n,sim_parameters)
) %>%
  ggplot(aes(x = synthetic)) +
  geom_histogram(
    aes(y = ..density..),
    color = "black",
    fill = "lightgray",
    linewidth = 1
    ) +
  geom_density(
    color = "red",
    linetype = "dashed",
    linewidth = 1
    ) +
  labs(
    x = "Sampled values",
    y = "Density",
    title = glue::glue("Empirical distribution of sampled values from a normal mixture")
  )
```

## Note for later

The text mentions that mixtures of binomial distributions may run into identifiability issues, which might be a problem because that is the application I have in mind!
